### 2010-10-11 papers/BMD2010/PhysicalParametersPaper/paper.tex
paper started

\documentclass{article}

\usepackage[lmargin=1in,rmargin=1in, tmargin=1in, bmargin=1in]{geometry}
\usepackage[colorlinks=true, linkcolor=blue]{hyperref}
\usepackage{graphicx}
\usepackage{eurosans}
\usepackage{subfiles}

\author{Jason K. Moore}
\title{Accurate Measurement of Eight Bicycles' Physical Parameters}

@@

\maketitle
\tableofcontents

\section{Introduction}
This work is intended to document the indirect measurement of eight real
bicycles' physical parameters. The physical parameters measured are those that
are needed for the benchmark bicycle model~\cite{Meijaard2007}. The work is
based off of the techiques used to measure the instrumented bicycle in
\cite{Kooijman2006} and \cite{Kooijman2008}. I used the same techniques as
Kooijman to measure a second instrumented bicycle~\cite{Moore2009a}. This
method had low accuracy and the $I_{yy}$ moments of inertia were
not measured for the fork and frame. Furthermore, very little data exists on
the physical parameters of different types of bicycles and this work aims to
provide a small sample of bicycles.

D\"{o}hring measured the physical parameters of a scooter~\cite{Dohring1953}.
Singh and Goel measured the physical parameters of a scooter~\cite{Singh1971}.
Roland and Massing measured the physical parameters of a bicycle in much the
same was as presented, including calculations of uncertainty from the indirect
measurement techniques~\cite{Roland1971}. Patterson used a swing to measure the
inertia of a bicycle and rider~\cite{Patterson2004}. This work is based off of the work
done by Kooijman~\cite{Kooijman2006}. I worked with Jodi Kooijman for a year and
was able to use much of the same apparatus and refine the measurement
technique.
\section{Bicycle Descriptions}
\subsection{Yellow Bicycle}
The yellow is a bicycle used in the lab to demonstrate that a bicycle is stable
at certain speeds. It is an aluminum road frame of unknown make with the
majority of components removed. The wheels, drop handlebar, seat, seat post and
bottom bracket are the only parts on the bike. The fork has been reversed to
''increase`` the stability of the bicycle. The bicycle was measured with both
the fork in normal position and the fork reversed.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{images/yellow.jpg}
    \end{center}
    \caption{Side view of the yellow bicycle with the fork in the reversed
    posistion.}
    \label{fig:YellowBike}
\end{figure}
\begin{table}[htbp]
	\begin{tabular}{ll}
		Style & Road\\
		Size & \\
		Wheels & 700c Aluminum Rims\\
		Tires & 28 x 1 3/8 x 1 5/8\\
		Sprocket & 6 speed free wheel\\
		Handlebars & Drop\\
	\end{tabular}
	\caption{Specifications of the Yellow Bicycle}
	\label{yellowBikeSpecs}
\end{table}
\subsection{Batavus Browser}
The Batavus Browser is lower priced Dutch city bike. We measured the physical
proprieties of the stock Browser model and also the same bike with the
instrumentation used in our experiments.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{images/browser.jpg}
    \end{center}
    \caption{Side view of the Batavus Browser as from the factory.}
    \label{fig:browser}
\end{figure}
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{images/browserIns.jpg}
    \end{center}
    \caption{The Batavus Browser with instrumentation as used
    in~\cite{Kooijman2009}}
    \label{fig:browserIns}
\end{figure}
\begin{table}[htbp]
	\begin{tabular}{ll}
        Price          & euro symbol 449\\
        Wheel Size     & 28''\\
        Weight         & 18.7 kg\\
        Frame Size     & 54\\
        Frame material & High Tensile Steel\\
        Fork material  & Steel\\
        Rims           & Aluminum Airline\\
        Spokes         & Stainless steel\\
        Tires          & Trax Highway, 28'' x 1 5/8 x 1 3/8\\
        Handlebar      & Steel painted\\
        Handles        & Batavus Comfort\\
        Saddle         & Selle Royal 8274\\
        Dynamo         & Basta Trio\\
        Headlight      & Batavus Logic Light\\
        Rearlight      & Battery light\\
        Pedals         & Batavus plastic antislip\\
        Final          & Trelock RS42\\
        Theft chip     & Present\\
        Crank          & Aluminum\\
        Rack           & Steel\\
        Quick Binder   & Binder Trio\\
        Mudguards      & Steel\\
        Bell           & Kenlight\\
	\end{tabular}
	\caption{2008 Batavus Browser Specifications.}
	\label{browserSpecs}
\end{table}

\subsection{2007 Bianchi Pista}
The Pista is a modern steel track bicycle. Specifications, taken from the Bianchi
website~\cite{Bianchi2009}, are as follows:
        \includegraphics[width=4in]{images/pista.jpg}

@@

    \caption{The Bianchi Pista hanging from a spring scale.}
    \label{fig:pista}
	\begin{tabular}{ll}
        Style            & Track Bike\\
        Size             & 57cm\\
        Color            & Gang Green\\
        Frame            & Bianchi DB CrMo, rear entry track dropouts\\
        Fork             & CrMo\\
        Retail Price     & \$579.99\\
        Headset          & VP AheadSet, 1''threadless\\
        Handlebar        & Bianchi steel track, 26.0mm\\
        Stem             & Bianchi alloy\\
        Brakes/Levers    & Shimano Tiagra front brake/lever\\
        Crankset         & Truvativ Touro 1.1, 48T\\
        Bottom Bracket   & Truvativ Power Spline cartridge\\
        Chain            & KMC\\
        Sprocket         & 16T fixed cog \& 16T Freewheel\\
        Pedals           & Campus pedals\\
        Wheels           & Bianchi Alex Solo wheelset, 28h rims\\
        Tires            & Continental UltraSport, 700x23C\\
        Saddle           & Bianchi Velo\\
        Seatpost         & Bianchi alloy, 27.2mm\\
        Size             & 57cm\\ 
        Seat Tube        & 570mm\\
        Top Tube-actual  & 560mm\\
        Top Tube-virtual & 560mm\\
        Chainstay        & 383mm\\
        Fork Rake        & 28mm\\
        Head Tube Angle  & 74.5$^{\circ}$\\
        Seat Tube Angle  & 75.5$^{\circ}$\\
        Wheelbase        & 967mm\\
        Standover Height & 32''  
	\caption{}
	\label{pistaSpecs}
\subsection{Gary Fisher Mountain Bike}
The Gary Fisher Ziggurat is a modern front suspended mountain bike.
        \includegraphics[width=4in]{images/fisher.jpg}
    \caption{The Gary Fisher mountain bike hanging from a spring scale.}
    \label{fig:fisher}
    \begin{tabular}{ll}
        Geometry & Genisis\\
        Frame & Aluminum, Damping, AC/DC, Carbon, Composite\\
        Front Wheel & DT Swiss XR4.1 Ceramic 559 17\\
        Tires & Michelin 26'' X 2.00 (5.2-559)\\
        Rear Wheel & DT Swiss 240s\\
        Rear hub & 9 speed\\
        Rear derailleur & SRAM X-0\\
        Crank & Bonterager 3-speed\\
        Seat post & Ritchey\\
        Fork & SID Rock Shox
    \caption{Specifications of the Gary Fisher Mountain Bike}
    \label{tab:gary}
\subsection{Batavus Crescendo Deluxe}
The Batavus Crescendo Deluxe is a contemporary of all-round cycling in cool
black and white. A great bike for a good end to touring. And that works fine
with 8 gears, suspension fork and suspension seatpost.
        \includegraphics[width=4in]{images/crescendo.jpg}
    \caption{A side view of the Batavus Crescendo Deluxe.}
    \label{fig:crescendo}
	\begin{tabular}{ll}
		Price             & from € 869\\
		Wheel Size        & 28''\\
		Weight            & 21.5 kg\\
		Frame Size        & 53\\
		Performance/speed & Shimano Nexus 8 speed with roller brakes\\
		Color             & White / black\\
		Brakes            & Shimano roller brakes (BR-IM50)\\
		Frame type        & Ladies mono tube frame Gentlemen trend\\
		Frame material    & Aluminum 7005\\
		Front Fork        & spring, Batavus\\
		Rims              & Aluminum Rodi VR19\\
		Spokes            & Stainless steel (adapted spoke pattern)\\
		Tires             & CST End\\
		Handlebar         & Steel aluminum look\\
		Stem              & Manually adjustable, Batavus Ergo Matic III\\
		Handles           & adjustable Batavus Ergo Grip II\\
		Saddle            & Selle Royal 5160\\
		Seatpost          & spring, Post Modern Compact\\
		Dynamo            & Shimano Hub Dynamo\\
		Headlight         & Trelock Hi-Power automatic halogen\\
		Rear              & Smart battery automatically\\
		Pedals            & aluminum antislip\\
		Final							& AXA Defender\\
		Chip theft        & Present\\
		Crank             & Aluminum\\
		Rack              & Aluminum\\
		Quick Binder      & Binder Quattro\\
		Mudguards         & Batavus hyalite II\\
		Bell              & Bibia Touring\\
		Balhoofdset       & VP semi-integrated, oversized\\
		Shifters          & Batavus (8-speed)\\
		Saddlebag         & Batavus saddlebag with plakset\\
	\caption{}
	\label{crescendoSpecs}
\subsection{Batavus Stratos Deluxe}
        \includegraphics[width=4in]{images/stratos.jpg}
    \caption{A side view of the Batavus Stratos Deluxe}
    \label{fig:stratos}
	\begin{tabular}{ll}
		Switching           & Spin System (shifter)\\
		Monochromatic frame & Black\\
		Chain guard         & Semi-open\\
		Mudguards           & Aluminum\\
		Rack                & Yes\\
		Extra drive         & Regardless Motori degree\\
		Type                & Men's\\
		Frame Material      & Aluminum\\
		Rim Material        & Aluminum\\
		Rear braking system & Roller brake\\
		Front braking system & Roller Brake\\
		Number of gears      & 7 Gears\\
		Gear System          & Shimano Nexus\\
		Wheel Size           & 28 inch\\
		Tirewidth            & 37 mm\\
		Price                & 749,00\\
	\caption{}
	\label{stratosSpecs}
\section{Accuracy}
Error propagation theory was utilized to calculate what the accuracy the
benchmark ideal parameters are based on the measurements taken. If $x$ is a
random variable and is defined as $x=f(u,v,\ldots)$ then the sample variance of
$x$ is defined as
    s_x^2 = \frac{1}{N-1}\sum^N_{i=1}
    \left[(u_i - \bar{u})^2\left(\frac{\partial x}{\partial u}\right)^2 +
    (v_i - \bar{v})^2\left(\frac{\partial x}{\partial v}\right)^2 +
    2(u_i - \bar{u})(v_i - \bar{v})\left(\frac{\partial x}{\partial u}\right)\left(\frac{\partial x}{\partial v}\right)
    + \ldots\right]
    \label{eqn:sampleVariance}
Using the definitions for variance and covariance
Equation~\ref{eqn:sampleVariance} can be simplified to
    s_x^2 = s_u^2\left(\frac{\partial x}{\partial u}\right)^2 +
            s_v^2\left(\frac{\partial x}{\partial v}\right)^2 +
            2s_{uv}\left(\frac{\partial x}{\partial u}\right)\left(\frac{\partial x}{\partial v}\right)
            + \ldots
    \label{eqn:variance}
If $u$ and $v$ are uncorrelated then $s_{uv}=0$. We assume that the all the
random variables hereafter are uncorrelated. Equation~\ref{eqn:variance} can be
used to calculated the variance of all types of functions. The most basic are
functions such as addtion
\begin{eqnarray}
    \label{eqn:addition}
    x &=&  au + bv\\
    s_x &=& a^2s_u^2 + b^2s_v^2
\end{eqnarray}

@@

\subsection{Wheel radius}
The radius of the front $r_\mathrm{F}$ and rear $r_\mathrm{R}$ wheels were
measured by measuring the linear distance traversed along the ground through
either 13 or 14 rotations of the wheel. Each wheel was measured separately and
the measurements were taken with a 72kg rider seated on the bicycle. A 30 meter
tape measure (resolution: 2mm) was taped on a flat level smooth floor. The tire
was marked with chalk and aligned with the tape measure. The accuracy of the
distance measurement is approximately $\pm0.005$m. The tires were pumped to the
recommended inflation pressure before the measurements. The wheel radius is
calculated by
\begin{equation}
	r=\frac{\textrm{distance}}{2\cdot\pi\cdot\textrm{rotations}}
	\label{eq:wheelRadius}
\end{equation}
\begin{equation}
    \sigma_r^{2}=\left(\frac{\sigma_d}{2\pi n}\right)^{2}
    \label{eq:WheelRadVar}
\end{equation}
\begin{figure}[tb]
	\begin{center}
		\includegraphics[width=4in]{images/tireChalk.jpg}
	\end{center}
	\caption{Wheel and tire with chalk mark aligned to the tape measure.}
	\label{fig:tireChalk}
\end{figure}
\begin{table}
	\begin{tabular}{llllllll}
	Bicycle   & B      & B*     & C      & G      & P      & S      & Y \& Y*\\
    \hline
	Front\\
	Pressure  & 3.5    & 3.5    & 4      & 4      & 6.9    & 4.5    & 4      \\
	Distance  & 28.060 & 27.980 & 27.980 & 29.044 & 29.366 & 27.772 & 27.925 \\
	Rotations & 13     & 13     & 13     & 14     & 14     & 13     & 13     \\
	Radius    & 0.344  & 0.343  & 0.343  & 0.330  & 0.334  & 0.340  & 0.342  \\
	Rear\\
	Pressure  & 3.5    & 3.5    & 4      & 4      & 6.9    & 4.5    & 3.5    \\
	Distance  & 27.850 & 27.835 & 27.768 & 29.788 & 29.212 & 29.774 & 27.884 \\
	Rotations & 13     & 13     & 13     & 14     & 14     & 14     & 13     \\
	Radius    & 0.341  & 0.341  & 0.340  & 0.339  & 0.332  & 0.338  & 0.341
	\end{tabular}
	\caption{}
	\label{tab:wheelRadius}
\end{table}
\subsection{Head tube angle}
The head tube angle was measured directly using an electronic level with a
$\pm0.2^{\circ}$ accuracy. The bicycle was fixed vertically, the front wheel
was aligned with the rear frame and the bicycle was unloaded. The steer axis
tilt $\lambda$ is the complement to the head tube angle.
\begin{equation}
    \lambda=\frac{\pi}{180^{\circ}}(90^{\circ}-\lambda_{ht})
\label{eq:headTubeAngle}
\end{equation}
\begin{equation}
    \sigma_\lambda^{2}=-\sigma_{\lambda_{ht}}^{2}\left(\frac{\pi}{180^\circ}\right)^2
    \label{eq:HeadTubeAngleVar}
\end{equation}
\begin{table}
	\begin{tabular}{lllllllll}
    \hline
	Bicycle   & B      & B*     & C      & G      & P      & S      & Y      & Y*\\
	Head tube angle & 67.1 & 67.1 & 69.0 & 71.1 & 74.2 & 73.1 & 72.7 & 70.6\\
	Steer axis tilt & 0.400 & 0.400 & 0.367 & 0.330 & 0.276 & 0.295 & 0.302 & 0.339
	\end{tabular}
	\caption{Head tube angle and steer axis tilt}
	\label{tab:lambda}
\end{table}

@@

        \includegraphics[width=4in]{images/massScale.jpg}

@@

        \includegraphics[width=4in]{images/pendDist.jpg}

@@

		\includegraphics[width=4in]{images/level.jpg}

@@

		\includegraphics[width=4in]{figures/angles.pdf}

@@

        \includegraphics{figures/triangle.pdf}

@@

        \includegraphics[width=4in]{../data/pendDat/graphs/BrowserFrameCompoundFirst1.png}

@@

        \includegraphics[width=4in]{../data/pendDat/graphs/CrescendoForkTorsionalFirst2.png}

@@

The period for a damped oscillation is
\begin{equation}
    T = \frac{2\pi}{\sqrt{1-\zeta^2}\omega_n}
    \label{eqn:periodDamped}
\end{equation}
and for light damping
\begin{equation}
    T \approx \frac{2\pi}{\omega_n}
    \label{eqn:period}
\end{equation}
The uncertainty in the period, $T$, can be determined from the fit. Firstly,
the variance of the fit is
\begin{equation}
    \sigma_y^2 =
    \frac{1}{N-5}\sum_{i=1}^N(y_{mi}-\bar{y}_m)^2-(y_{pi}-\bar{y}_m)^2
    \label{eqn:fitVariance}
\end{equation}
The covariance matrix of the fit function can be formed
\begin{equation}
    \mathbf{U} = \sigma_y^2\mathbf{H}^{-1}
    \label{eqn:covariance}
\end{equation}
where $\mathbf{H}$ is the Hessian~\cite{Hubbard1989b}. $\mathbf{U}$ is a $5\times5$ matrix with the
variances of each of the five fit parameters along the diagonal. This assumes
that the parameters are uncorrelated [note for Jason: you may have to take into
account correlation for this..maybe checkout the R function for nls]. The
variance of $T$ can be computed using the variance of $\zeta$ and $\omega$. It
is important to note that the uncertainties in the period are very low
($<10e-4$), even for the fits with low $r^2$ values due to the beating issue.
\subsection{Torsional Pendulum}
A torsional pendulum was used to measure all of moments of inertia about axes
in the lateral symmetric plane of each of the wheels, fork and frame. The
pendulum is made up of a rigid mount, an upper clamp, a torsion rod, and
various lower clamps.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{images/fixture.jpg}
    \end{center}
    \caption{The rigid pendulumfixture mounted to a concrete column.}
    \label{fig:fixture}
\end{figure}
A 5mm diameter, 1m long mild steel rod was used as the torsion
spring. A lightweight, low relative moment of inertia clamp was constructed
that could clamp the rim and the tire. The wheel was hung freely such that the
center of mass aligned with the torsional pendulum axis and then it was
secured. The wheel was then perturbed and it oscillated about the pendulum axis.
The rate gyro was mounted on the clamp in line with the pendulum axis. The
radial moment of inertia can can calculated as such:

The torsional pendulum was calibrated using a known moment of inertia. 
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{images/rod.jpg}
    \end{center}
    \caption{The steel calibration rod.}
    \label{fig:rod}
\end{figure}
A torsional pendulum almost identical to the one used in
\cite{Kooijman2006} was used to measure the averaged period $\overline{T}_i$ of
oscillation of the rear frame at three different
orientation angles $\beta_i$, where $i=1$, $2$, $3$, as shown in
Fig.~\ref{fig:triangle}. The parts were perturbed lightly, less than 1 degree,
and allowed to oscillate about the pendulum axis through at least ten periods.
The time of oscillation was recorded via a stopwatch ($\pm1$~s). This was done
three times for each frame and the recorded times were averaged. The
coefficient of elasticity $k$ for the torsional pendulum had previously been
measured in \cite{Kooijman2006} and found to be $k=5.01\pm0.01$
$\frac{\textrm{Nm}}{\textrm{rad}}$.

>>>>

\documentclass{bmd2010p}

%\usepackage[colorlinks=true, linkcolor=blue]{hyperref}
\usepackage{graphicx}
\usepackage{subfiles}

@@

\begin{center}
\fontsize{14}{20}{\bf Accurate Measurement of Bicycle Parameter[3pt]}
\end{center}

%%%%%%%%%%%%%%%% authors %%%%%%%%%%%%%%%
\begin{center}
\normalsize{\bf{Jason K. Moore$^{*}$, Mont Hubbard, A. L. Schwab$^\#$,
            J. D. G. Kooijman$^\dag$}}
\end{center}

\begin{center}
\begin{tabular}{c}
$^*$ Mechanical and Aerospace Engineering\\
University of California, Davis\\
One Shields Avenue, Davis, CA, 95616, USA\\
e-mail: jkmoor@ucdavis.edu, mhubbard@ucdavis.edu\\
\end{tabular}
\begin{tabular}{c}
$^\#$ Laboratory for Engineering Mechanics\\
Delft University of Technology\\
Mekelweg 2, 2628CD Delft, The Netherlands\\
e-mail: a.l.schwab@tudelft.nl, jodikooijman@gmail.com\\
\end{tabular} \\ \vspace{2.5ex}

@@

\section*{ABSTRACT}
Accurate measurements of a bicycle's physical parameters are required for
realistic dynamic simulations and analysis. For the most basic models the
geometry, mass, mass location and mass distributions must be measured. More complex models
require measurements of tire characteristics, human characteristics, friction, stiffness, damping, etc. This
paper concerns the measurement of the minimal bicycle parameters required for
the benchmark bicycle presented in~\cite{Meijaard2007}. This
model is composed of four rigid bodies, has ideal rolling and frictionless joints,
and is assumed to be laterally symmetric. A set of 25
parameters is used to describe the geometry, mass, mass location and
mass distribution of each of the rigid bodies. The experimental methods
described herein are based primarily on the work
done in~\cite{Kooijman2006} but have been refined for improved accuracy and
methodology. Koojiman's work was preceded by \cite{Roland1971} who measured a bicycle in a
similar fashion and both~\cite{Dohring1953} and~\cite{Singh1971} who have used
similar techniques with scooters. We measured the characteristics of six
different bicycles, two of which were set up in two different configurations.
This is a total of eight different parameter sets that can be used with, but not
limited to, the benchmark bicycle model. The accuracy of all the measurements
are presented up through the eigenvalue prediction of the linear model and are
based on error propagation theory with correlations taken into account.

The six bicycles, chosen for both variety and convenience, are as follows:
\emph{Batavus Browser}, a Dutch style city bicycle measured with and without
instrumentation as described in~\cite{Kooijman2009a}; \emph{Batavus Stratos
Deluxe}, a Dutch style sporty city bicycle; \emph{Batavus Crescendo Deluxe} a
Dutch style city bicycle with a suspended fork; \emph{Gary Fisher Mountain
Bike}, a hardtail mountain bicycle; \emph{Bianchi Pista}, a modern steel frame
track racing bicycle; and \emph{Yellow Bicycle}, a stripped down aluminum frame
road bicycle measured in two configurations, the second with the fork rotated
in the headtube 180 degrees for larger trail.
\begin{keywords}
bicycle,
parameters,
eigenvalues,
Bode.
\end{keywords}

\section{INTRODUCTION}

This work is intended to document the indirect measurement of eight real
bicycles' physical parameters. The physical parameters measured are those that
are needed for the benchmark bicycle model~\cite{Meijaard2007}. The work is
based off of the techiques used to measure the instrumented bicycle in
\cite{Kooijman2006} and \cite{Kooijman2008}. I used the same techniques as
Kooijman to measure a second instrumented bicycle~\cite{Moore2009a}. This
method had low accuracy and the $I_{yy}$ moments of inertia were
not measured for the fork and frame. Furthermore, very little data exists on
the physical parameters of different types of bicycles and this work aims to
provide a small sample of bicycles.

D\"{o}hring measured the physical parameters of a scooter~\cite{Dohring1953}.
Singh and Goel measured the physical parameters of a scooter~\cite{Singh1971}.
Roland and Massing measured the physical parameters of a bicycle in much the
same was as presented, including calculations of uncertainty from the indirect
measurement techniques~\cite{Roland1971}. Patterson used a swing to measure the
inertia of a bicycle and rider~\cite{Patterson2004}. This work is based off of the work
done by Kooijman~\cite{Kooijman2006}. I worked with Jodi Kooijman for a year and
was able to use much of the same apparatus and refine the measurement
technique.

\section{Bicycle Descriptions}
\subsection{Yellow Bicycle}
The yellow is a bicycle used in the lab to demonstrate that a bicycle is stable
at certain speeds. It is an aluminum road frame of unknown make with the
majority of components removed. The wheels, drop handlebar, seat, seat post and
bottom bracket are the only parts on the bike. The fork has been reversed to
''increase`` the stability of the bicycle. The bicycle was measured with both
the fork in normal position and the fork reversed.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/yellow.jpg}
    \end{center}
    \caption{Side view of the yellow bicycle with the fork in the reversed
    posistion.}
    \label{fig:YellowBike}
\end{figure}
\subsection{Batavus Browser}
The Batavus Browser is lower priced Dutch city bike. We measured the physical
proprieties of the stock Browser model and also the same bike with the
instrumentation used in our experiments.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/browser.jpg}
    \end{center}
    \caption{Side view of the Batavus Browser as from the factory.}
    \label{fig:browser}
\end{figure}
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/browserIns.jpg}
    \end{center}
    \caption{The Batavus Browser with instrumentation as used
    in~\cite{Kooijman2009}}
    \label{fig:browserIns}
\end{figure}

\subsection{2007 Bianchi Pista}
The Pista is a modern steel track bicycle. Specifications, taken from the Bianchi
website~\cite{Bianchi2009}, are as follows:
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/pista.jpg}
    \end{center}
    \caption{The Bianchi Pista hanging from a spring scale.}
    \label{fig:pista}
\end{figure}
\subsection{Gary Fisher Mountain Bike}
The Gary Fisher Ziggurat is a modern front suspended mountain bike.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/fisher.jpg}
    \end{center}
    \caption{The Gary Fisher mountain bike hanging from a spring scale.}
    \label{fig:fisher}
\end{figure}
\subsection{Batavus Crescendo Deluxe}
The Batavus Crescendo Deluxe is a contemporary of all-round cycling in cool
black and white. A great bike for a good end to touring. And that works fine
with 8 gears, suspension fork and suspension seatpost.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/crescendo.jpg}
    \end{center}
    \caption{A side view of the Batavus Crescendo Deluxe.}
    \label{fig:crescendo}
\end{figure}
\subsection{Batavus Stratos Deluxe}
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/stratos.jpg}
    \end{center}
    \caption{A side view of the Batavus Stratos Deluxe}
    \label{fig:stratos}
\end{figure}
\section{Accuracy}
Error propagation theory was utilized to calculate what the accuracy the
benchmark ideal parameters are based on the measurements taken. If $x$ is a
random variable and is defined as $x=f(u,v,\ldots)$ then the sample variance of
$x$ is defined as
\begin{equation}
    s_x^2 = \frac{1}{N-1}\sum^N_{i=1}
    \left[(u_i - \bar{u})^2\left(\frac{\partial x}{\partial u}\right)^2 +
    (v_i - \bar{v})^2\left(\frac{\partial x}{\partial v}\right)^2 +
    2(u_i - \bar{u})(v_i - \bar{v})\left(\frac{\partial x}{\partial u}\right)\left(\frac{\partial x}{\partial v}\right)
    + \ldots\right]
    \label{eqn:sampleVariance}
\end{equation}
Using the definitions for variance and covariance
Equation~\ref{eqn:sampleVariance} can be simplified to
\begin{equation}
    s_x^2 = s_u^2\left(\frac{\partial x}{\partial u}\right)^2 +
            s_v^2\left(\frac{\partial x}{\partial v}\right)^2 +
            2s_{uv}\left(\frac{\partial x}{\partial u}\right)\left(\frac{\partial x}{\partial v}\right)
            + \ldots
    \label{eqn:variance}
\end{equation}
If $u$ and $v$ are uncorrelated then $s_{uv}=0$. We assume that the all the
random variables hereafter are uncorrelated. Equation~\ref{eqn:variance} can be
used to calculated the variance of all types of functions. The most basic are
functions such as addtion
\begin{eqnarray}
    \label{eqn:addition}
    x &=&  au + bv\\
    s_x &=& a^2s_u^2 + b^2s_v^2
\end{eqnarray}

@@

\subsection{Wheel radius}
The radius of the front $r_\mathrm{F}$ and rear $r_\mathrm{R}$ wheels were
measured by measuring the linear distance traversed along the ground through
either 13 or 14 rotations of the wheel. Each wheel was measured separately and
the measurements were taken with a 72kg rider seated on the bicycle. A 30 meter
tape measure (resolution: 2mm) was taped on a flat level smooth floor. The tire
was marked with chalk and aligned with the tape measure. The accuracy of the
distance measurement is approximately $\pm0.005$m. The tires were pumped to the
recommended inflation pressure before the measurements. The wheel radius is
calculated by
\begin{equation}
	r=\frac{\textrm{distance}}{2\cdot\pi\cdot\textrm{rotations}}
	\label{eq:wheelRadius}
\end{equation}
\begin{equation}
    \sigma_r^{2}=\left(\frac{\sigma_d}{2\pi n}\right)^{2}
    \label{eq:WheelRadVar}
\end{equation}
\begin{figure}[tb]
	\begin{center}
		\includegraphics[width=4in]{../../../images/tireChalk.jpg}
	\end{center}
	\caption{Wheel and tire with chalk mark aligned to the tape measure.}
	\label{fig:tireChalk}
\end{figure}
\begin{table}
	\begin{tabular}{llllllll}
	Bicycle   & B      & B*     & C      & G      & P      & S      & Y \& Y*\\
    \hline
	Front\\
	Pressure  & 3.5    & 3.5    & 4      & 4      & 6.9    & 4.5    & 4      \\
	Distance  & 28.060 & 27.980 & 27.980 & 29.044 & 29.366 & 27.772 & 27.925 \\
	Rotations & 13     & 13     & 13     & 14     & 14     & 13     & 13     \\
	Radius    & 0.344  & 0.343  & 0.343  & 0.330  & 0.334  & 0.340  & 0.342  \\
	Rear\\
	Pressure  & 3.5    & 3.5    & 4      & 4      & 6.9    & 4.5    & 3.5    \\
	Distance  & 27.850 & 27.835 & 27.768 & 29.788 & 29.212 & 29.774 & 27.884 \\
	Rotations & 13     & 13     & 13     & 14     & 14     & 14     & 13     \\
	Radius    & 0.341  & 0.341  & 0.340  & 0.339  & 0.332  & 0.338  & 0.341
	\end{tabular}
	\caption{}
	\label{tab:wheelRadius}
\end{table}
\subsection{Head tube angle}
The head tube angle was measured directly using an electronic level with a
$\pm0.2^{\circ}$ accuracy. The bicycle was fixed vertically, the front wheel
was aligned with the rear frame and the bicycle was unloaded. The steer axis
tilt $\lambda$ is the complement to the head tube angle.
\begin{equation}
    \lambda=\frac{\pi}{180^{\circ}}(90^{\circ}-\lambda_{ht})
\label{eq:headTubeAngle}
\end{equation}
\begin{equation}
    \sigma_\lambda^{2}=-\sigma_{\lambda_{ht}}^{2}\left(\frac{\pi}{180^\circ}\right)^2
    \label{eq:HeadTubeAngleVar}
\end{equation}
\begin{table}
	\begin{tabular}{lllllllll}
    \hline
	Bicycle   & B      & B*     & C      & G      & P      & S      & Y      & Y*\\
	Head tube angle & 67.1 & 67.1 & 69.0 & 71.1 & 74.2 & 73.1 & 72.7 & 70.6\\
	Steer axis tilt & 0.400 & 0.400 & 0.367 & 0.330 & 0.276 & 0.295 & 0.302 & 0.339
	\end{tabular}
	\caption{Head tube angle and steer axis tilt}
	\label{tab:lambda}
\end{table}

@@

        \includegraphics[width=4in]{../../../images/massScale.jpg}

@@

        \includegraphics[width=4in]{../../../images/pendDist.jpg}

@@

		\includegraphics[width=4in]{../../../images/level.jpg}

@@

		\includegraphics[width=4in]{../../../figures/angles.pdf}

@@

        \includegraphics{../../../figures/triangle.pdf}

@@

        \includegraphics[width=4in]{../../../plots/PendFit/BrowserFrameCompoundFirst1.png}

@@

        \includegraphics[width=4in]{../../../plots/PendFit/CrescendoForkTorsionalFirst2.png}

@@

The period for a damped oscillation is
\begin{equation}
    T = \frac{2\pi}{\sqrt{1-\zeta^2}\omega_n}
    \label{eqn:periodDamped}
\end{equation}
and for light damping
\begin{equation}
    T \approx \frac{2\pi}{\omega_n}
    \label{eqn:period}
\end{equation}
The uncertainty in the period, $T$, can be determined from the fit. Firstly,
the variance of the fit is
\begin{equation}
    \sigma_y^2 =
    \frac{1}{N-5}\sum_{i=1}^N(y_{mi}-\bar{y}_m)^2-(y_{pi}-\bar{y}_m)^2
    \label{eqn:fitVariance}
\end{equation}
The covariance matrix of the fit function can be formed
\begin{equation}
    \mathbf{U} = \sigma_y^2\mathbf{H}^{-1}
    \label{eqn:covariance}
\end{equation}
where $\mathbf{H}$ is the Hessian~\cite{Hubbard1989b}. $\mathbf{U}$ is a $5\times5$ matrix with the
variances of each of the five fit parameters along the diagonal. This assumes
that the parameters are uncorrelated [note for Jason: you may have to take into
account correlation for this..maybe checkout the R function for nls]. The
variance of $T$ can be computed using the variance of $\zeta$ and $\omega$. It
is important to note that the uncertainties in the period are very low
($<10e-4$), even for the fits with low $r^2$ values due to the beating issue.
\subsection{Torsional Pendulum}
A torsional pendulum was used to measure all of moments of inertia about axes
in the lateral symmetric plane of each of the wheels, fork and frame. The
pendulum is made up of a rigid mount, an upper clamp, a torsion rod, and
various lower clamps.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/fixture.jpg}
    \end{center}
    \caption{The rigid pendulumfixture mounted to a concrete column.}
    \label{fig:fixture}
\end{figure}
A 5mm diameter, 1m long mild steel rod was used as the torsion
spring. A lightweight, low relative moment of inertia clamp was constructed
that could clamp the rim and the tire. The wheel was hung freely such that the
center of mass aligned with the torsional pendulum axis and then it was
secured. The wheel was then perturbed and it oscillated about the pendulum axis.
The rate gyro was mounted on the clamp in line with the pendulum axis. The
radial moment of inertia can can calculated as such:

The torsional pendulum was calibrated using a known moment of inertia. 
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/rod.jpg}
    \end{center}
    \caption{The steel calibration rod.}
    \label{fig:rod}
\end{figure}
A torsional pendulum almost identical to the one used in
\cite{Kooijman2006} was used to measure the averaged period $\overline{T}_i$ of
oscillation of the rear frame at three different
orientation angles $\beta_i$, where $i=1$, $2$, $3$, as shown in
Fig.~\ref{fig:triangle}. The parts were perturbed lightly, less than 1 degree,
and allowed to oscillate about the pendulum axis through at least ten periods.
The time of oscillation was recorded via a stopwatch ($\pm1$~s). This was done
three times for each frame and the recorded times were averaged. The
coefficient of elasticity $k$ for the torsional pendulum had previously been
measured in \cite{Kooijman2006} and found to be $k=5.01\pm0.01$
$\frac{\textrm{Nm}}{\textrm{rad}}$.
